---
title: 'Lab1'
author: 'liume102@student.liu.se'
date: '12 Nov 2024'
output:
  html_document:
    number_sections: true
    toc: true
    fig_width: 7
    fig_height: 4.5
    theme: readable
    highlight: tango
---

# Introduction

This is the first lab in the Machine Learning In this lab, contains the following tasks:1. Handwritten digit recognition with Knearest neighbors.2.Linear regression and ridge regression.3. Logistic regression and basis function expansion.4. Theory

# Handwritten digit recognition with Knearest neighbors



## Load and check data

```{r, message = FALSE}
# Load packages
library('ggplot2') # visualization
library('ggthemes') # visualization
library('scales') # visualization
library('dplyr') # data manipulation
library('mice') # imputation
library('randomForest') # classification algorithm
library('caret')
```

Now that our packages are loaded and we divide it into training, validation and test sets (50%/25%/25%)
```{r, message=FALSE, warning=FALSE}
digitals <- read.csv('../data/optdigits.csv', stringsAsFactors = F,header = FALSE)

train_index <- createDataPartition(digitals$V65, p = 0.5, list = F) 
train_digitals <- digitals[train_index,]
remainingData <- digitals[-train_index, ]
validationIndex <- createDataPartition(remainingData$V65, p = 0.5, list = FALSE)
valid_digitals <- remainingData[validationIndex, ]
test_digitals <- remainingData[-validationIndex, ]
```

# KNN to fit classification model
```{r, message=FALSE, warning=FALSE}
library(knn)



```
